% dvips -t letter hw_math.dvi -o hw_math.ps ; ps2pdf hw_math.ps
\documentclass[11pt,titlepage,fleqn]{article}

\input{hw626_header}

\renewcommand{\baselinestretch}{1.1}

%--------------------------------------------------------------
\begin{document}
%-------------------------------------------------------------

\begin{spacing}{1.2}
\centering
{\large \bf Problem Set 2: Linear algebra and vectors [math]} \\
GEOS 626: Applied Seismology, Carl Tape \\
Assigned: January 29, 2018 --- Due: February 5, 2018 \\
Last compiled: \today \\
\end{spacing}

%------------------------

\subsection*{Overview}

The purpose of this problem set is to review a few mathematical concepts that are important to seismology. These concepts will be useful in following the class reading \citep{SteinWysession,ShearerE2} and for future homework assignments.

Except for Problem 1, you are welcome to use Matlab to support your answers. If you find yourself doing horrendous algebra, then probably you have made a prior mistake.

%------------------------

\subsection*{Problem 1 (5.0). Matrix decompositions}

This problem must be done by hand, unless specified. ``By hand'' means showing your steps and not using decimal notation; for example, write the expression $(4 +\sqrt{3})/5$ rather than 1.1464.
Be as clear as possible in explaining what equations you are using.
%You are welcome to check your answers using Matlab.

\medskip\noindent
The matrix for this problem is
%
\begin{equation*}
\bA =  \left[ \begin{array}{rrr}
    -1  &   0  &  -4 \\
     0  &   2  &   1 \\
     1  &   1  &   4 \\
\end{array} \right]
\end{equation*}

\input{hw_math_input}

%------------------------

\subsection*{Problem 2 (3.0). Matrix operations}

\begin{enumerate}

\item (0.3) Determine whether the following three vectors are linearly dependent:
%
\begin{eqnarray*}
\left[ \begin{array}{r} 1 \\ -2 \\ 4  \end{array} \right],
\left[ \begin{array}{r} 0 \\ 1 \\ 1 \end{array} \right],
\left[ \begin{array}{r} 1 \\ 4 \\ -2 \end{array} \right]
\end{eqnarray*}

%----------------

\item (0.3) Find the matrix $\bA$ ($T(\bx) = \bA\bx$) of the following linear transformations:
%
\begin{enumerate}
\item $\mathbb{R}^2 \rightarrow \mathbb{R}^2$ reflection about $y = 0$.
\item $\mathbb{R}^2 \rightarrow \mathbb{R}^2$ reflection about $x = 0$.
\item $\mathbb{R}^3 \rightarrow \mathbb{R}^3$ reflection through $xy$-plane.
\end{enumerate}
%
Hint: What does $\bA$ do to the input vector $(x,y)$?

%----------------

\item (0.3)
%You can use Matlab for any tedious steps.

\begin{enumerate}
\item (0.1) Find the matrix, $\bA$, ($T(\bx) = \bA\bx$) of the linear transformation described by
%
\begin{equation*}
T \left( \left[ \begin{array}{r} 1 \\ 0 \\ 0  \end{array} \right] \right) = \left[ \begin{array}{r} 1 \\ 2 \end{array} \right],
\hspace{0.5cm}
T \left( \left[ \begin{array}{r} 0 \\ 1 \\ 0  \end{array} \right] \right) = \left[ \begin{array}{r} 2 \\ 4 \end{array} \right],
\hspace{0.5cm}
T \left( \left[ \begin{array}{r} 0 \\ 0 \\ 1  \end{array} \right] \right) = \left[ \begin{array}{r} 3 \\ 6 \end{array} \right]
\end{equation*}

\item (0.2) Find the matrix, $\bA$, ($T(\bx) = \bA\bx$) of the linear transformation described by
%
\begin{equation*}
T \left( \left[ \begin{array}{r} 1 \\ 0 \\ 0  \end{array} \right] \right) = \left[ \begin{array}{r} 1 \\ 1 \\ 0 \end{array} \right],
\hspace{0.5cm}
T \left( \left[ \begin{array}{r} 2 \\ 1 \\ 0  \end{array} \right] \right) = \left[ \begin{array}{r} 1 \\ 0 \\ -1 \end{array} \right],
\hspace{0.5cm}
T \left( \left[ \begin{array}{r} 3 \\ 2 \\ 1  \end{array} \right] \right) = \left[ \begin{array}{r} 0 \\ 1 \\ 1 \end{array} \right]
\end{equation*}

\end{enumerate}

%----------------

\pagebreak
\item (0.8) 
\begin{enumerate}
\item (0.2) Compute the angle between $\bu = (1,1,1)^T$ and $\bv = (1,2,3)^T$.

\item Compute the area of the parallelogram defined by these vectors using three different methods:
\begin{enumerate}
\item (0.2) Use the height vector, $\bh = \bv - \proj_{\ssL} \bv$, where $\ssL$ is the line in the direction of $\bu$ (\refFig{fig:proj}).
%Sketch the various vectors and the parallelogram.
\item (0.2) Use the cross product formula, $\| \bu \times \bv \|$ (note: {\em not} the version $\|\bu\|\|\bv\|\sin\theta$, which is used in part a).
\item (0.2) Compute $\sqrt{\det(\bA^T \bA)}$, where $\bA = [\;\bu \; \bv\;]$.
\end{enumerate}

\end{enumerate}

%----------------

\item (0.3) \citep[][p.~473, P10]{SteinWysession}.

Prove that the magnitude of a vector is preserved by an orthogonal transformation.

Hint: Let $A\bu = \bv$ and show that $\|\bu\| = \|\bv\|$.

%----------------

\item (0.6) Show that 
%
\begin{eqnarray}
\hlf \left( \bd - \bG\bem \right)^T \left( \bd - \bG\bem \right)
&=& \hlf \bem^T\bA\bem + \bb^T\bem + c
\label{lsq1}
\end{eqnarray}
%
where
%
\begin{equation}
\bA = \bG^T\bG
\,,\hspace{20pt}
\bb = -\bG^T\bd
\,,\hspace{20pt}
c = \hlf\bd^T\bd
\,.
\end{equation}
%
See Note \footnote{The standard construction of a least squares problem is often given by
%
\begin{equation}
\bG\bem = \bd,
\label{lsq}
\end{equation}
%
which has solution
%
\begin{equation}
\bem = \left( \bG^T\bG \right)^{-1} \bG^T \bd.
\label{lsq_sol}
\end{equation}
%
This formulation in \refEq{lsq} is equivalent to minimizing the function
%
\begin{equation}
F(\bem) = \hlf \sum_{i=1}^N \left(\;d_i - \left[ \bG\bem \right]_i \;\right)^2
\;=\; \hlf \left\| \bd  - \bG\bem \right\|_{L2}^2
\;=\; \hlf \left( \bd  - \bG\bem \right)^T \left( \bd - \bG\bem \right)
\end{equation}
}
for context for this problem.

Hint: For a scalar $a$, note that $a = a^T$.

%----------------

\item (0.4) Find all $\bx$ such that $\bA\bx = \bb$, where $\bb = (1, 1, 0)^T$ and
%
\begin{eqnarray*}
\bA =  \left[ \begin{array}{rrrr}
     0  &   0  &   1 & -4 \\
     1  &   -2  &   0 & -3 \\
     -1  &   2  &   1 & -1 \\
\end{array} \right].
\end{eqnarray*}
%
Express the solution as a sum of three vectors (including free parameters).

Hint: Feel free to use a computer to help.
%Matlab is permissible (though it is not needed).

\end{enumerate}

%------------------------

\pagebreak
\subsection*{Problem 3 (1.0). Div, grad, curl, and all that}

\citep[][p.~473, P13]{SteinWysession} \\
Consider the vector field
%
\begin{equation}
\bu(x,y,z) = \bxh(3x^2y^2 + z) + \byh(2x^3y+2y) + \bzh(x).
\label{vec}
\end{equation}
%
(Note: Make sure you write down \refEq{vec} correctly; otherwise your answers will all be incorrect!) Calculate the following four expressions:
%
\begin{enumerate}
\item (0.4) $\bdel \cdot \bu$
\item (0.2) $\bdel \times \bu$
\item (0.2) $\nabla^2\bu$
\item (0.2) A scalar field $\phi(x,y,z)$ such that $\bu=\bdel\phi$
\end{enumerate}

%------------------------

%\pagebreak
\subsection*{Problem 4 (1.0). Taylor series}

\begin{enumerate}
\item (0.5) Derive the Taylor series expansion of $f(x) = \ln(x)$ about the point $x = x_0$. List terms up to the third order.

\item (0.5) Let $x_0 = 3.5$. In all of the plots, use $x$ limits from $-1$ to 10 (\verb+xlim([-1 10])+).
%
\begin{enumerate}
\item (0.2) Plot the zeroth-, first-, second-, and third-order {\bf terms} on a single plot.

\item (0.2) Plot the zeroth-, first-, second-, and third-order {\bf Taylor series} on a single plot. Also plot $\ln(x)$.
(The $n$th-order Taylor series is the sum up to the $n$th-order term.) 

\item (0.1) Normalized residuals can be expressed as $\ln(A/B)$, where both $A > 0$ and $B > 0$.
The same can be applied to functions such as $f(x) > 0$ and $g(x) > 0$.
Consider the normalized residuals $\ln[ f(x) / g_k(x) ]$ where $f(x) = \ln(x)$ and $g_k(x)$ is the $k$th-order Taylor series approximation. What is the approximate range of the normalized residual over the range $x = [3,4]$?
\end{enumerate}

\end{enumerate}

%------------------------

%\pagebreak
\subsection*{Problem} \howmuchtime\

%-------------------------------------------------------------
\bibliographystyle{agu08}
\bibliography{carl_abbrev,carl_main,carl_source,carl_him,carl_alaska}
%-------------------------------------------------------------

\clearpage\pagebreak
\input{notes_matrix_figs}

%-------------------------------------------------------------
\end{document}
%-------------------------------------------------------------
